---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ulrb

<!-- badges: start -->
  [![R-CMD-check](https://github.com/pascoalf/ulrb/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/pascoalf/ulrb/actions/workflows/R-CMD-check.yaml)
  [![test-coverage](https://github.com/pascoalf/ulrb/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/pascoalf/ulrb/actions/workflows/test-coverage.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/ulrb)](https://CRAN.R-project.org/package=ulrb)
[![GPLv3 license](https://img.shields.io/badge/License-GPLv3-blue.svg)](http://perso.crans.org/besson/LICENSE.html)
[![Github All Releases](https://img.shields.io/github/downloads/pascoalf/ulrb/total.svg)]()
`r lifecycle::badge("stable")`
<!-- badges: end -->

The R package **ulrb** stands for **Unsupervised Machine Learning definition of the Rare Biosphere**. As the name suggests, it applies unsupervised learning principles to define the rare biosphere.

More specifically, the partitioning around medoids (k-medoids) algorithm is 
used to divide phylogenetic units (ASVs, OTUs, Species, ...) within a 
microbial community (usually, a sample) into clusters. The clusters are then
ordered based on a user-defined classification vector. By default, our 
method classifies all phylogenetic units in one of these: "rare", "undetermined"
or "abundant". However, this can be changed by the user if necessary.

For detailed theory behind our reasoning for this definition of the microbial 
rare biosphere, results and applications, see our paper Pascoal et al., 2023 (in preparation). 
For more details on the R functions used and data wrangling please see the package 
documentation. 

For tutorials and documentation of the **urlb** package, visite our website: link.

## Installation

You can install the development version of ulrb from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("pascoalf/ulrb")
```

We are working on a CRAN publication. We will then provide instructions for CRAN installation.

## Example

This is a basic example which shows you how to use ulrb to divide phylogenetic units into three classifications (rare, undetermined and abundant):

```{r example}
library(ulrb)
## basic example
define_rb(nice_tidy)
```

With ulrb, you can also format your original species table, get an automatic number of clusters and plot the results:

```{r, fig.width= 10}
# nice is an OTU table in wide format
head(nice)

# first, we tidy the "nice" OTU table
sample_names <- c("ERR2044662", "ERR2044663", "ERR2044664",
                   "ERR2044665", "ERR2044666", "ERR2044667",
                   "ERR2044668", "ERR2044669", "ERR2044670")

# If data is in wide format, with samples in cols
nice_tidy <- prepare_tidy_data(nice,
                               sample_names = sample_names,
                               samples_in = "cols")

# second, we apply ulrb algorithm in automatic setting
nice_classification_results <- define_rb(nice_tidy, automatic = TRUE)

# third, we plot microbial community and the quality of k-medoids clustering
plot_ulrb(nice_classification_results, taxa_col = "OTU", plot_all = TRUE)

# In case you want to inspect the result of a particular sample, do:
plot_ulrb(nice_classification_results, taxa_col = "OTU", sample_id = "ERR2044662")

```
